Stirring Up Neural ODEs: A Chemical Engineer's perspective

### 2025-03-20 talk.tex
finally done with the talk

\title[NODEs]{Stirring Up Neural ODEs: A Chemical Engineer's Recipe for Dynamic Modeling
}

@@

\date{ME255NN Neural ordinary differential equations  \\
\today}

@@

\begin{frame}{ResNets and RNNs}
    \begin{block}{A state-space representation: Euler discretized system}
    \begin{align*}
        x^+ &= x + f_{NN} (x, u; \theta) &\qquad x^+ = f_{NN} (x, u; \theta)\notag \\ 
        y &= g_{NN} (x, u; \theta) &\qquad y = g_{NN} (x, u; \theta)
        \label{eq:resrnn}
    \end{align*}
    Parallels with FIR and ARX models in system identification \footnote[frame]{\cite{ljung:1999}}.\\
    Backprop strategy: Brute-force chain rule (store intermediate states).
    \end{block}
    \begin{columns}
    {\begin{column}{0.6\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth, page=2]{ABC_plot.pdf}
    \end{figure}
\end{column}}
{\begin{column}{0.3\textwidth}
    \ If $x\in\R^n, \theta\in R^p$, $S = \dfrac{\partial x}{\partial \theta^T}\in\R^{n\times \textcolor{red}{p}}$
\end{column}}
\end{columns}
\end{frame}

\begin{frame}{Don't want to store intermediate states? Forward sensitivities for IVPs \footnote{Derivation in \cite{rawlings:ekerdt:2020}}}
    \begin{itemize}
    \item Consider a non-linear state-space model ($y=x$):
    \begin{align*}
        \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)
    \end{align*}
    \item Sensitivity is defined as and evolves as \footnote{Using Clairaut's Theorem on Mixed Partials}:
    \begin{align*}
        S &= \dfrac{\partial x}{\partial \theta^T} \\
        \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
        \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
        \dfrac{\partial }{\partial \theta^T} f
    \end{align*}
    \item Augmented system of ODEs:
    \begin{align*}
        \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
        \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
        \qquad
        \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
        \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
    \end{align*}
    \item Notice again: $S\in \R^{n\times p}$ i.e sensitivities we need to solve
    increase \textcolor{red}{linearly} with $p$. Imagine how bad this could get for FNN with millions of parameters and 
    states!
    \end{itemize}
\end{frame}

\begin{frame}{How to avoid the linear cost? Pontryagin's adjoint method}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{oldies.png}\\
        \textit{One mastered control, one shaped space, one gambled with certainty}.\\
    \end{figure}
\end{frame}

\begin{frame}{Backward adjoint method}
    \begin{itemize}
        \item Consider following optimization problem and the Lagrange functional:
        \begin{align*}
            \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \hat{x}}^2 dt = \int_0^t g(x)dt \notag \\
            \min\limits_{\theta, \lambda}\mathcal{L} &= \int_0^t g dt + \int_0^t \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
        \end{align*}
        \item The augmented system of ODEs is \footnote{Refer \cite{sengupta:friston:penny:2014} for paper and 
        \url{https://youtu.be/k6s2G5MZv-I?si=ZIyP7MvgzTK7D9kn} for a walkthrough. Additionally, 
        I have worked out the derivation step-by-step in my report.}:
        \[
        \begin{bmatrix} \dot{x} \\ \dot{\lambda^T} \end{bmatrix} = 
        \begin{bmatrix} f_{NN} \\ -\lambda^T \dfrac{\partial f_{NN}}{\partial x^T}\end{bmatrix}
        \qquad
        \begin{bmatrix} x(t) \\ \lambda^T(t) \end{bmatrix} =
        \begin{bmatrix} x_t \\ - \dfrac{\partial g}{\partial x^T}|_{t} \end{bmatrix}
        \label{eq:augsens}
        \]
        \item The gradient of loss wrt parameters is calculated as:
        \[
            \dfrac{\partial V}{\partial \theta^T} = \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
        \]
    \end{itemize}
\end{frame}

\begin{frame}{An ultimate panacea?}
    \begin{columns}
    {\begin{column}{0.45\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=\textwidth, page=4]{ABC_plot.pdf}
    \end{figure}
    \end{column}}
    {\begin{column}{0.5\textwidth}
    \begin{block}{}
        \begin{itemize}
            \item Consider $\dot{\lambda} = \pm 10 \lambda$ and $\lambda(0) = 0$
            \item We see that the adjoint variable will either explode or vanish.
            This heavily impacts the learning problem. 
            \item The Jacobian $\dfrac{\partial f}{\partial x^T}$ can be ill-conditioned.
            \item We still solve the adjoint equation by quadrature, thus numerical noise 
            can still be a problem.
        \end{itemize}
    \end{block}
    \begin{block}{Recent remedies}
        \begin{itemize}
            \item Don't solve adjoint over long horizons, but rather 
            small pieces and store the initial conditions for those intermediate
            pieces.
            \item Popularly called as \textit{checkpointing} \footnote[frame]{\cite{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}}.
        \end{itemize}
    \end{block}
    \end{column}}
    \end{columns}
\end{frame}

@@

\begin{frame}{NODE}
    \begin{figure}
        \centering
        \includegraphics[width=0.7\textwidth, page=1]{ABC_plot.pdf}\\
        Concentration profiles of species A, B, and C in a batch reactor. The
    NODE model also does quite a good job to predict the concentration profiles. 
    In fact for such simple problem, we hardly see any difference between ResNet and NODE.
    \end{figure}
\end{frame}

\begin{frame}{Generative latent model using CNF (Generative=Probabilistic)}
\end{frame}

>>>>

\title[NODEs]{Stirring Up Neural ODEs: A Chemical Engineer's perspective}

@@

\date{ME255NN Report  \\
\today}

@@

\begin{frame}{ResNets and RNNs}
    \begin{block}{A state-space representation: Euler discretized system}
    \begin{align*}
        x^+ &= x + f_{NN} (x, u; \theta) &\qquad x^+ = f_{NN} (x, u; \theta)\notag \\ 
        y &= g_{NN} (x, u; \theta) &\qquad y = g_{NN} (x, u; \theta)
        \label{eq:resrnn}
    \end{align*}
    Parallels with FIR and ARX models in system identification \footnote[frame]{\cite{ljung:1999}}.\\
    Backprop strategy: Brute-force chain rule (store intermediate states).
    \end{block}
    \begin{columns}
    {\begin{column}{0.6\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth, page=2]{ABC_plot.pdf}
    \end{figure}
\end{column}}
{\begin{column}{0.3\textwidth}
    \ If $x\in\R^n, \theta\in \R^p$, $S = \dfrac{\partial x}{\partial \theta^T}\in\R^{n\times \textcolor{red}{p}}$
\end{column}}
\end{columns}
\end{frame}

\begin{frame}{Don't want to store intermediate states? Forward sensitivities for IVPs \footnote{Derivation in \cite{rawlings:ekerdt:2020}}}
    \begin{itemize}
    \item Consider a non-linear state-space model ($y=x$):
    \begin{align*}
        \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)
    \end{align*}
    \item Sensitivity is defined as and evolves as \footnote{Using Clairaut's Theorem on Mixed Partials}:
    \begin{align*}
        S &= \dfrac{\partial x}{\partial \theta^T} \\
        \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
        \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
        \dfrac{\partial }{\partial \theta^T} f
    \end{align*}
    \item Augmented system of ODEs:
    \begin{align*}
        \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
        \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
        \qquad
        \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
        \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
    \end{align*}
    \item Notice again: $S\in \R^{n\times p}$ i.e sensitivities we need to solve
    for increase \textcolor{red}{linearly} with $p$. Imagine how bad this could get for FNN with millions of parameters and 
    states!
    \end{itemize}
\end{frame}

\begin{frame}{How to avoid the linear cost? Pontryagin's adjoint method}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{oldies.png}\\
        %\textit{One mastered control, one shaped space, one gambled with certainty}.\\
    \end{figure}
\end{frame}

\begin{frame}{Backward adjoint method}
    \begin{itemize}
        \item Consider following optimization problem and the Lagrange functional given data $\tilde{x}$:
        \begin{align*}
            \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \tilde{x}}^2 dt = \int_0^t g(x)dt \notag \\
            \min\limits_{\theta, \lambda}\mathcal{L} &= \int_0^t g dt + \int_0^t \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
        \end{align*}
        \item The augmented system of ODEs is \footnote{Refer \cite{sengupta:friston:penny:2014} for paper and 
        \url{https://youtu.be/k6s2G5MZv-I?si=ZIyP7MvgzTK7D9kn} for a walkthrough. Additionally, 
        I have worked out the derivation step-by-step in my report.}:
        \[
        \begin{bmatrix} \dot{x} \\ \dot{\lambda^T} \end{bmatrix} = 
        \begin{bmatrix} f_{NN} \\ -\lambda^T \dfrac{\partial f_{NN}}{\partial x^T}\end{bmatrix}
        \qquad
        \begin{bmatrix} x(t) \\ \lambda^T(t) \end{bmatrix} =
        \begin{bmatrix} x_t \\ - \dfrac{\partial g}{\partial x^T}|_{t} \end{bmatrix}
        \label{eq:augsens}
        \]
        \item The gradient of loss wrt parameters is calculated as:
        \[
            \dfrac{\partial V}{\partial \theta^T} = \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
        \]
    \end{itemize}
\end{frame}

\begin{frame}{An ultimate panacea?}
    \begin{columns}
    {\begin{column}{0.45\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=\textwidth, page=4]{ABC_plot.pdf}
    \end{figure}
    \end{column}}
    {\begin{column}{0.5\textwidth}
    \begin{block}{}
        \begin{itemize}
            \item Consider $\dot{\lambda} = \pm 10 \lambda$ and $\lambda(0) = 0$
            \item We see that the adjoint variable will either \textit{explode} or 
            \textit{vanish}.
            This heavily impacts the learning problem. 
            \item The Jacobian $\dfrac{\partial f}{\partial x^T}$ can be ill-conditioned.
            \item We solve the adjoint equation by quadrature, thus numerical noise 
            can still be a problem.
        \end{itemize}
    \end{block}
    \begin{block}{Remedies}
        \begin{itemize}
            \item Don't solve adjoint over long horizons, but rather 
            small windows and store the initial conditions for those intermediate
            windows.
            \item Popularly called as \textit{checkpointing} \footnote[frame]{\cite{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}}.
        \end{itemize}
    \end{block}
    \end{column}}
    \end{columns}
\end{frame}

@@

\begin{frame}{NODE}
    \begin{figure}
        \centering
        \includegraphics[width=0.65\textwidth, page=1]{ABC_plot.pdf}\\
        Concentration profiles of species A, B, and C in a batch reactor. The
    NODE model also does quite a good job to predict the concentration profiles. 
    In fact for such simple problem, we hardly see any difference between ResNet and NODE
    \footnote{All the case studies in this report are built using \texttt{Torchdiffeq} by \cite{torchdiffeq}}.
    \end{figure}
\end{frame}

\begin{frame}{Generative latent model using CNF}
    \begin{itemize}
        \item \textbf{Latent} = States in state-space model, since we usually cannot measure
        them directly. Thus, we construct state-estimators.
        \item \textbf{Generative latent} = State-space model that also propagates uncertainty
        in estimates i.e. you can sample and \textit{generate} multiple trajectories after learning.
        \item Surprise! Good old Kalman filter can be called a generative latent model!
        \footnote{Refer \cite{rawlings:mayne:diehl:2020} for derivation of the Kalman filter using
        Bayesian estimation principles.}
        \item Consider a simple linear transform:
        \begin{align*}
            x &\in \mathcal{N}(m, \sigma^2 I) \\
            y &= Ax \\
            y &\in \mathcal{N}(Am, A\sigma^2 A^T)
        \end{align*}
        \item Bayesian approach: $\textcolor{blue}{p(y|x)} \sim p(x|y) \textcolor{red}{p(y)}$.
        Assuming an \textcolor{red}{uniform prior} the likelihood problem is 
        \[
        \theta = \arg\max\limits_{\theta} \log p(x|y; \theta) \qquad p(x|y) = A^{-1} y \sim \mathcal{N}(m, \sigma^2I)        \]
        \item Ofcourse, $A$ is assumed to be non-singular (i.e. unique mapping between $x$ and $y$;
        \textit{bijective mapping}).
    \end{itemize}
\end{frame}

\begin{frame}{Extending the bayesian framework for any non-linear transform}
    \begin{itemize}
        \item Consider following series of transformations (\textit{planar-flow}):
        \[
        x^+ = f_{NN}(x, u; \theta) \qquad x(0) \sim \mathcal{N}(m, \sigma^2I) \qquad x(t) \sim ?
        \]
        \item Use \textit{change of variables}, compute the
        \textcolor{red}{determinant} of Jacobian ($\sim N(o^3)$) and conserve the probability mass as well
        as maximize the likelihood.
        \item Instead use continuous framework (\textit{continuous normalizing flow}) \footnote{Refer \cite{chen:rubanova:bettencourt:duvenaud:2018} for proof}:
        \begin{align*}
        \dfrac{dx}{dt} = f_{NN}(x, u; \theta) &\qquad x(0) \sim \mathcal{N}(m, \sigma^2I) \qquad x(t) \sim ?\\
        \dfrac{\partial p(x)}{\partial t} &=  -\mathrm{tr}\nabla_x f_{NN}
        \end{align*}
        \item Cost to get the trace of Jacobian is now linear $\sim N(o)$
    \end{itemize}
\end{frame}

\begin{frame}{Generative latent model for kinetics}
    \begin{figure}
        \centering
        \includegraphics[width=\textwidth, page=3]{ABC_plot.pdf}\\
    \end{figure}
    \begin{align*}
        \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
        = \begin{bmatrix}
            -k_{1}c_A + k_{-1}c_B  \\
            2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
            k_{2}c_B\\
            -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial c^T}\right)
        \end{bmatrix};
        &\qquad
        \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
        \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
        \min\limits_{\theta} V(c) &= -\log p(c) + \norm{c - \tilde{c}}^2
    \end{align*}
\end{frame}

\begin{frame}{Robustness analysis of NODEs \footnote{Refer \cite{yan:du:tan:feng:2019}}}
    \begin{figure}[h]
        \centering
        \includegraphics[width=0.5\textwidth, page=5]{ABC_plot.pdf} \\
        Robustness analysis of ODEs. We see that the solution $c_{B2}$ 
        is sandwiched between two perturbed solutions $c_{B1}, c_{B3}$, 
        though all approach the same steady-state. Also note that the slightly time-shifted
        solution appears as a perturbed solution \footnote{Refer \cite{ascher:petzold:1998}
        for well-posedness theorem of IVPs.}
        \label{fig:robust}
    \end{figure}
    \center \textbf{Theoretical bound}
    \begin{align*}
        \norm{x(t+T)-x(t)} \leq \norm{\int_t^{t+T} f_{NN}(x, u; \theta) dt}
    \end{align*} 
\end{frame}

\begin{frame}{How to train your NODE?}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.55\textwidth, page=6]{ABC_plot.pdf}  \\
    If we train based only on prediction loss, the model can overfit.
    However, if we penalize the Jacobian term ($\dfrac{\partial f_{NN}}{\partial x^T}$), we can regularize the model
    to damp the oscillations \footnote{\cite{finlay:jacobsen:nurbekyan:oberman:2020} \\ 
    Access the code for all figures at \url{https://github.com/dakeprithvi/2025a_cnf}}
    \end{figure}
\end{frame}
